---
title: "Beyond linear regression"
subtitle: "Ch4: Poisson Regression"
author: "Joe Skufca"
date: "2024-03-23"
output:
  html_notebook: default
---

This work take from Ch 4 of *Beyond Multiple Linear Regression* by Roback and Legler

https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html




## Packages

Recommended from the book:

```{r}
# Packages required for Chapter 4
library(gridExtra)  # grid graphics
library(knitr)   #knitting documents
library(kableExtra) # pretty tables
library(mosaic) # stats teaching utilities
library(xtable) # html tables
library(pscl) #poly sci computational lab
library(multcomp) 
library(pander) #pandoc writing
library(MASS) #modern applied statistics book
library(tidyverse)

```

Adding my own favorites.

```{r}

library(janitor)
library(here)
library(broom)

```


```{r}
data_github_url="https://raw.githubusercontent.com/proback/BeyondMLR/master/data/"
```



## Sec 4.2

If $\lambda_i$ is the average rate for explanatory condition $i$, then we will use

$$log (\lambda_i) = \beta_0+\beta_1 x_i$$


## Sec 4.4 Household size in Phillippines

I will read data directly from github:


```{r}
fHH1=read_csv(paste0(data_github_url,"fHH1.csv"))
```

```{r, nhouse, fig.align="center",out.width="60%", fig.cap='Distribution of household size in 5 Philippine regions.',echo=FALSE, warning=FALSE, message=FALSE}
ggplot(fHH1, aes(total)) + 
  geom_histogram(binwidth = .25, color = "black", 
                 fill = "white") + 
  xlab("Number in the house excluding head of household") +
  ylab("Count of households")
```

### Some EDA

```{r}
fHH1 %>% 
     mutate(agegroup=cut(age,breaks=20)) %>%
     summarize(n=n(),mean=mean(total),var=var(total),.by=agegroup) %>%
     arrange(agegroup)
```


```{r}
fHH1 %>% 
     mutate(agegroup=cut(age,breaks=20)) %>%
     summarize(n=n(),mean=mean(total),var=var(total),.by=age) %>%
     ggplot(aes(age,log(mean)))+
     geom_point()+
     geom_smooth()


```
by region?

```{r}
fHH1 %>% 
          summarize(n=n(),
               mean=mean(total),
               var=var(total),
               .by=c(age,location) )%>%
     ggplot(aes(age,log(mean),color=location))+
     geom_point()+
     geom_smooth()
```



#### Modeling

```{r}
modela = glm(total ~ age, family = poisson, data = fHH1)

summary(modela)
```

```{r}
confint(modela)
exp(confint(modela))
```

#### 4.4.4 Deviances

```{r}
model0 <- glm(total ~ 1, family = poisson, data = fHH1)
anova(model0, modela, test = "Chisq")
```

#### 4.4.6 Second order model


```{r}
fHH1 <- fHH1 %>% mutate(age2 = age*age)
modela2 = glm(total ~ age + age2, family = poisson, 
              data = fHH1)

summary(modela2)
```

```{r}
modela22 = glm(total ~ age + I(age^2)+location, family = poisson, 
              data = fHH1)

summary(modela22)
```

```{r}
modela23 = glm(total ~ poly(age,2), family = poisson, 
              data = fHH1)

summary(modela23)
```

```{r}
anova(modela, modela2, test = "Chisq")
```

####  4.4.7 Add a covariate

```{r}
modela2L = glm(total ~ age + age2 + location, 
               family = poisson, data = fHH1)

summary(modela2L)
```

```{r}
anova(modela,modela2, modela2L, test = "Chisq")
```

#### 4.4.8 deviance residuals

```{r}
augment(modela) %>% ggplot(aes(.fitted,.resid))+geom_point()+geom_smooth()
```


###4.6 campus crime

```{r}
c_data0=read_csv(paste0(data_github_url,"c_data.csv"))


```

```{r}
hist(c_data0$nv,breaks=20)
```

```{r}
c_data0 %>% tabyl(type,region) 
```
Data revison prior to building Table 4.4

```{r}
c.data=c_data0 %>%
     mutate(type=factor(type),
            region=factor(region) %>% fct_recode(S="SW",S="SE")) %>%
     filter(nvrate<5.3)


```

```{r}
c.data %>% 
     summarize(across(nv:nvrate,list(mean=mean,var=var)),
               .by = c(region,type)) %>%
     arrange(region,type)
```


### 4.8 models

```{r}
modeltr <- glm(nv ~ type + region+type:region, family = poisson,
               offset = log(enroll1000), data = c.data)

summary(modeltr)
```



