---
title: "Chap 8 exercise"
subtitle: "MLR"
author: "Joe Skufca"
date:  "2024-03-06"
output: html_notebook
---

Practice problems from Chapter 8 has only one problem:

## 8.1. Treadmill data analysis 
The original research goal for the treadmill data set used for practice problems
in the last two chapters was to replace the costly treadmill oxygen test with a cheap to find running time
measurement but there were actually quite a few variables measured when the run time was found – maybe
we can replace the treadmill test result with a combined prediction built using a few variables using the MLR
techniques. The following code will get us re-started in this situation.

## Packages that we will typically load


Here are three of my favorites that I always load, +GGally.


```{r}
library(tidyverse)
library(janitor)
library(here)
library(gridExtra)
library(broom)

library(GGally) #for data ease
library(car)
library(effects)
```


## Read Data

> Since I do not have to follow code from the book, I choose to use my standard formatting for variables.
> As such, I juse janaitor to rename my variables.

```{r}
treadmill <- read_csv("http://www.math.montana.edu/courses/s217/documents/treadmill.csv") |>
     clean_names()
```

> The book has us compute a first linear model called `tm1`

```{r}
tm1 <- lm(tread_mill_ox ~ run_time, data = treadmill)
summary(tm1)
```


### 8.1.1. 
Fit the MLR that also includes the running pulse (RunPulse), the resting pulse (RestPulse), body
weight (BodyWeight), and Age (Age) of the subjects. Report and interpret the R2
for this model.

```{r}
tm2=lm(tread_mill_ox ~ run_time+
            run_pulse+rest_pulse+body_weight+age,
       data=treadmill)
summary(tm2)
```
The resultant $R^2$ value is

```{r}
glance(tm2)$r.squared
```

### 8.1.2. 
Compare the R2 and the adjusted R2
to the results for the SLR model that just had RunTime in the
model. What do these results suggest?

```{r}
glance(tm1)$adj.r.squared
glance(tm2)$adj.r.squared
```

The MLR explains more of the variances, even when correcting fot the additional complexity of the model.  

We note that it may still not be the *best* model, as there may be a smaller model that does better.  

### 8.1.3. 

Interpret the estimated RunTime slope coefficients from the SLR model and this MLR model. Explain
the differences in the estimates.

```{r}
tm1$coefficients
tm2$coefficients
```

* In the SLR, for a one minute increase in run_time we estimate that the mean tread_mill_ox will decrease by 3.3 units.
* In the MLR, for a one minute increase in run_time, we estimate a decrease of 2.7 units in oxygen, *after controlling for the other variables in the model (run_pulse, rest_pulse, body_weight, and age).


### 8.1.4. 
Find the VIFs for this model and discuss whether there is an issue with multicollinearity noted in these
results.
```{r}
vif(tm2)
```

Although there is some light colineartiy, the vif values for all variables are below 1.6,
indicating that we should have no siginficant concern about colinearity.

### 8.1.5. 

Report the value for the overall F-test for the MLR model and interpret the result.


```{r}
summary(tm2)
```


With an overall F-statistic of 22.41 and a p-value of 1.7e-8,  there is strong evidence against the null 
hypothesis that *no explanatory varialbes should be included*.   


### 8.1.6. 
Drop the variable with the largest p-value in the MLR model and re-fit it. Compare the resulting R2
and adjusted R2 values to the others found previously.

```{r}
tm3=lm(tread_mill_ox ~ run_time+
            run_pulse+body_weight+age,
       data=treadmill)
summary(tm3)
```

Dropping rest_pulse, we find that adjusted R-sqared increases from .7811 to .7883 .


### 8.1.7. 
Use the dredge function as follows to consider some other potential reduced models and report the top
two models according to adjusted R2 values. What model had the highest R2.
Also discuss and compare
the model selection results provided by the delta AICs here

```{r}
library(MuMIn)
options(na.action = "na.fail") #Must run this code once to use dredge
md=dredge(tm2, rank = "AIC",
       extra = c("R^2",
                 "adjR^2",
                 myadjRsq = function(x) summary(x)$adj.r.squared
                 )) %>% as_tibble() %>% clean_names()

md
```

The top two models (by adjusted R^2 are):

```{r}
md %>% slice_max(n=2, order_by = myadj_rsq)
```

> Note that the adj R^2 as computed by MuMIn is different than computed by lm.

The top models exludes both body_weigh and rest_pusle, while the second best excludes just rest_pulse.


These selected modes are the same top two as indicated by AIC.

### 8.1.8. 
For one of the models, interpret the Age slope coefficient. Remember that only male subjects between
38 and 57 participated in this study. Discuss how this might have impacted the results found as compared to
a more general population that could have been sampled from.



The age coefficient for the top model indicates that for a 1 year increase in age,
treadmill_ox will decrease by about 0.26, controlling for run_pulse and run_time.

As all the the subjects are likely to be "past their prime" (in the stages of phycial decline with age),
the indicated performance decline is not unexpected.  

Had we included a broader spectrum of aged subjects (say 18-60), we might reasonably find that 
performance increases with age for younger men, while decreasing with age for older men.

### 8.1.9. 
The following code creates a new three-level variable grouping the ages into low, middle, and high for
those observed. The scatterplot lets you explore whether the relationship between treadmill oxygen and run
time might differ across the age groups.

```{r}
treadmill2 <- treadmill %>% mutate(ageb = factor(cut(age, breaks = c(37, 44.5, 50.5, 58))))
summary(treadmill2$ageb)
treadmill2 %>%
     ggplot(mapping = aes(
          x = run_time,
          y = tread_mill_ox,
          color = ageb,
          shape = ageb
     )) +
     geom_point(size = 1.5, alpha = 0.5) +
     geom_smooth(method = "lm") +
     theme_bw() +
     scale_color_viridis_d(end = 0.8) 

#+
#     facet_grid(rows = vars(ageb))
```

### 8.1.10. 
Fit the MLR that contains a RunTime by Ageb interaction – do not include any other variables.
Compare the R2 and adjusted R2
results to previous models.

```{r}
tm4=lm(tread_mill_ox ~run_time*ageb,data=treadmill2)

summary(tm4)
glance(tm4)
```

```{r}
glance(tm1)
```

```{r}
tm5=lm(tread_mill_ox ~run_time+age,data=treadmill2)
glance(tm5)
```

Based on both R-squared and adjusted R-sq, we see that the model which adds age categories
performs better than the SLR model without age, but also better than MLR with both runtime and age.


I will also look at the interaction model (using age) to see if it helps:

```{r}
tm6=lm(tread_mill_ox ~run_time*age,data=treadmill2)
glance(tm6)
```

Interestingly, this model does not perform as well as the age categorized model.

### 8.1.11. 
Find and report the results for the F-test that assesses evidence relative to the need for different
slope coefficients.

```{r}
Anova(tm4)
```

Performing an Anova calculation on `tm4`, we see that the F-test associated with the
interaction (run_time:ageb) results in a p-value of .3904, showing little evidence to
reject the null hypothesis that all slopes are the same.

### 8.1.13. 
Fit the additive model with RunTime and predict the mean treadmill oxygen values for subjects with
run times of 11 minutes in each of the three Ageb groups.

```{r}
tm4_add=lm(tread_mill_ox ~run_time+ageb,data=treadmill2)

summary(tm4)
```
```{r}
newdata=tibble(run_time=11,ageb=levels(treadmill2$ageb))
predict(tm4_add,newdata)
```

### 8.1.14. 
Find the F-test results for the binned age variable in the additive model. Report and interpret those
results.

```{r}
Anova(tm4_add)
```

There is weak evidence against the null-hypothesis that the intercept is the same for all three groups.

We note that based on model performance (as a assed my adjusted R^2) we would use separate intercepts.

