---
title: "Day 15 and 16 class"
subtitle: "MLR"
author: "Joe Skufca"
date:  "2024-02-11"
output: html_notebook
---

Materials from Chapter 8.

## Packages that we will typically load


Here are three of my favorites that I always load, +GGally.


```{r}
library(tidyverse)
library(janitor)
library(here)
library(gridExtra)
library(broom)

library(GGally) #for data ease
library(car)
```


## Read Data

```{r}
snotel_s <- read_csv("http://www.math.montana.edu/courses/s217/documents/snotel_s.csv")
```

```{r message=FALSE, warning=FALSE}
snotel_s %>% ggpairs(columns = c(4:6,3)) +
theme_bw()
```


```{r}
m1 <- lm(Snow.Depth ~ Elevation, data = snotel_s)
m2 <- lm(Snow.Depth ~ Min.Temp, data = snotel_s)
m3 <- lm(Snow.Depth ~ Max.Temp, data = snotel_s)
library(effects)
plot(allEffects(m1, residuals = T), main = "SLR: Effect of Elevation")
plot(allEffects(m2, residuals = T), main = "SLR: Effect of Min Temp")
plot(allEffects(m3, residuals = T), main = "SLR: Effect of Max Temp")
```

```{r}
summary(m1)
summary(m2)
summary(m3)
```



```{r}
m4 <- lm(Snow.Depth ~ Elevation + Min.Temp + Max.Temp, data = snotel_s)
summary(m4)
```

```{r}
plot(allEffects(m4, residuals = T), main = "MLR model with Elev, Min, & Max Temps")

```


## 8.2 validity

```{r}
par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(m4, pch = 16, sub.caption = "")
title(main="Diagnostics for m4", outer=TRUE)
```

```{r}
snotel_s %>% slice(9,10)
```

```{r}
predict(m4,snotel_s %>% slice(9,10))
```

```{r}
library(broom)
tidy(m4)
```

```{r}
augment(m4)
```

```{r}
m5 <- lm(Snow.Depth ~ Elevation + Min.Temp + Max.Temp, data = snotel_s %>% slice(-9))
summary(m5)
```

```{r}
plot(allEffects(m5, residuals = T), main = "MLR model with NE Ent. Removed")
```

```{r}
par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(m5, pch = 16, sub.caption = "")
title(main="Diagnostics for m5", outer=TRUE)
```

```{r}
augment(m5)

```


```{r}
m6 <- lm(Snow.Depth ~ Elevation + Min.Temp + Max.Temp, data = snotel_s %>% slice(-c(9,22)))
summary(m6)
par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(m6, pch = 16, sub.caption = "")
title(main="Diagnostics for m6", outer=TRUE)
```

```{r}
plot(allEffects(m6, residuals = T), main = "MLR model with n = 23")
```

```{r}
set.seed(307)
snotel_final <- snotel_s %>% slice(-c(9,22))
snotel_final <- snotel_final %>%
#Creates first and second set of simulated responses
mutate(SimulatedSnow = simulate(m6)[[1]],
SimulatedSnow2 = simulate(m6)[[1]]
)



```

```{r}
r1 <- snotel_final %>% ggplot(aes(x = Elevation, y = Snow.Depth)) +
geom_point() +
theme_bw() +
labs(title = "Real Responses")

r2 <- snotel_final %>% ggplot(aes(x = Max.Temp, y = Snow.Depth)) +
geom_point() +
theme_bw() +
labs(title = "Real Responses")
r3 <- snotel_final %>% ggplot(aes(x = Min.Temp, y = Snow.Depth)) +
geom_point() +
theme_bw() +
labs(title = "Real Responses")

s1 <- snotel_final %>% ggplot(aes(x = Elevation, y = SimulatedSnow)) +
geom_point(col = "forestgreen") +
theme_bw() +
labs(title = "First Simulated Responses")

s2 <- snotel_final %>% ggplot(aes(x = Max.Temp, y = SimulatedSnow)) +
geom_point(col = "forestgreen") +
theme_bw() +
labs(title = "First Simulated Responses")
s3 <- snotel_final %>% ggplot(aes(x = Min.Temp, y = SimulatedSnow)) +
geom_point(col = "forestgreen") +
theme_bw() +
labs(title = "First Simulated Responses")
s12 <- snotel_final %>% ggplot(aes(x = Elevation, y = SimulatedSnow2)) +
geom_point(col = "skyblue") +
theme_bw() +
labs(title = "Second Simulated Responses")
s22 <- snotel_final %>% ggplot(aes(x = Max.Temp, y = SimulatedSnow2)) +
geom_point(col = "skyblue") +
theme_bw() +
labs(title = "Second Simulated Responses")
s32 <- snotel_final %>% ggplot(aes(x = Min.Temp, y = SimulatedSnow2)) +
geom_point(col = "skyblue") +
theme_bw() +
labs(title = "Second Simulated Responses")

grid.arrange(r1, r2, r3, s1, s2, s3, s12, s22, s32, ncol = 3)
```

## 8.3 Interpretation

```{r}
summary(m6)
```
For a 1 [units of $x_k$] increase in $x_k$, we estimate the mean of $y$ to change by $b_k$ [units of y],
after controlling for [list of other explanatory variables in model].
To make this more concrete, we can recreate some points in the Elevation term-plot. To do this, we first
need the mean of the “other” predictors, Min.Temp and Max.Temp.


#### consider simulated values / diagnostic plots

```{r}
m6s=lm(SimulatedSnow~Elevation+Min.Temp+Max.Temp,data=snotel_final);

par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(m6s, pch = 16, sub.caption = "")
title(main="Diagnostics for first simulated responses", outer=TRUE)

```

```{r}
m6s2=lm(SimulatedSnow2~Elevation+Min.Temp+Max.Temp,data=snotel_final);

par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(m6s2, pch = 16, sub.caption = "")
title(main="Diagnostics for first simulated responses2", outer=TRUE)

```


### following the book



```{r}
mean(snotel_final$Min.Temp)
mean(snotel_final$Max.Temp)
```

Lets create a tibble like was done on page 321, but allowing all computations in R.

```{r}
my_effects_tibble=tibble(Elevation=c(5000,6000,8000),
                         Min.Temp=mean(snotel_final$Min.Temp),
                         Max.Temp=mean(snotel_final$Max.Temp)) %>%
     augment(m6,newdata=.)

my_effects_tibble %>% ggplot(aes(x=Elevation,y=.fitted))+
     geom_point(size=2)+geom_line(col="tomato")+
     theme_bw()+
     ggtitle("Custom effects plot")
```

At more points:

```{r}
elevs <- seq(from = 5000, to = 8000, length.out = 30)
newdata1 <- tibble(Elevation = elevs, Min.Temp = 27.826,
Max.Temp = 36.3913)
newdata1
```
>pg 322   


```{r}

#predict(m6, newdata = newdata1, interval = "confidence") %>% head(10) 

augment(m6,newdata = newdata1, interval = "confidence") %>% 
     ggplot(aes(x=Elevation,y=.fitted,ymin=.lower,ymax=.upper)) +
     geom_line(col="tomato")+
     geom_ribbon(fill="green",alpha=.2)+
     theme_bw()+
     ggtitle("My effects plot")

```

> Note that there are a variety of ways to
> note that each term in an MLR is only a particular value given the other variables in the model. We can
> use words such as “holding the other variables constant” or “after adjusting for the other variables” or “in
> a model with. . . ” or “for observations with similar values of the other variables but a difference of 1 unit
> in the predictor..”. The main point is to find words that reflect that this single slope coefficient might be
> different if we had a different overall model and the only way to interpret it is conditional on the other model
> components.

## 8.4 comparing models


## 8.5 General recommendations:



```{r}
library(corrplot)
par(mfrow = c(1,1), oma = c(0,0,1,0))
corrplot.mixed(cor(snotel_s %>% slice(-c(9,22)) %>% select(3:6)),
upper.col = c(1, "orange"), lower.col = c(1, "orange"))
round(cor(snotel_s %>% slice(-c(9,22)) %>% select(3:6)), 2)
```

#### variable inflation factor

```{r}
library(car)
vif(m6)
sqrt(vif(m6))
```

```{r}
elev1 <- lm(Elevation ~ Min.Temp + Max.Temp, data = snotel_s %>% slice(-c(9,22)))
summary(elev1)
```



## 8.6 parameter inference using t-distribution

```{r}
m5 <- lm(Snow.Depth ~ Elevation + Max.Temp, data = snotel_s %>% slice(-c(9,22)))
summary(m5)
```


```{r}
confint(m5)
```


```{r}
sqrt(vif(m5))
```
### Bootstrap??

## 8.7 overall F-test

are any slope coefficients non-zero

## 8.8 Case study


6 steps:

1. Hypothesis
2. Plot data and assess validity
     * quantitative variables
     * independence
     * linearity (include partial residuals plot)
     * understand multicolinearity
     * equal variance
     * normality of residuals
     * no influential points
3. p-values
4. conclusions
5. size 
6. scope of inference

```{r}
library(openintro)
data(satgpa)
satgpa <- as_tibble(satgpa)
satgpa <- satgpa %>% rename(gender = sex , #Renaming variables
satv = sat_v, satm = sat_m, satsum = sat_sum,
hsgpa = hs_gpa,
fygpa = fy_gpa)
satgpa %>%
select(-4) %>%
ggpairs() +
theme_bw()
```


```{r}
gpa1 <- lm(fygpa ~ satv + satm, data = satgpa)
summary(gpa1)
glance(gpa1)
```

```{r}
par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(gpa1, sub.caption = "")
title(main="Diagnostics for GPA model with satv and satm", outer=TRUE)
```

```{r}
vif(gpa1)

sqrt(vif(gpa1))
```

```{r}
plot(allEffects(gpa1, residuals = T))
```

```{r}
confint(gpa1)
```
```{r}
library(effectsize)
```


```{r}
cohens_f_squared(gpa1)
```

#### real discriptions of size

```{r}
predict(gpa1, newdata = tibble(satv = 30,satm = 60), interval = "confidence")
predict(gpa1, newdata = tibble(satv = 30, satm = 60), interval = "prediction")
```

```{r}
# Remake effects plots with added 95% PIs
dv1 <- tibble(satv = seq(from = 24, to = 76, length.out = 50), satm = rep(54.4, 50))
mv1 <- as_tibble(predict(gpa1, newdata = dv1, interval = "confidence"))
pv1 <- as_tibble(predict(gpa1, newdata = dv1, interval = "prediction"))
mres_GPA_v <- bind_cols(dv1, mv1, pv1 %>% select(-fit))
# Rename CI and PI limits to have more explicit column names:
mres_GPA_v <- mres_GPA_v %>% rename(lwr_CI = lwr...4, upr_CI = upr...5,
lwr_PI = lwr...6, upr_PI = upr...7)
v1 <- mres_GPA_v %>% ggplot() +
geom_line(aes(x = satv, y = fit), lwd = 1) +
geom_ribbon(aes(x = satv, ymin = lwr_CI, ymax = upr_CI), alpha = .4,
fill = "beige", color = "darkred", lty = 2, lwd = 1) +
geom_ribbon(aes(x = satv, ymin = lwr_PI, ymax = upr_PI), alpha = .1,
fill = "gray80", color = "grey", lty = 3, lwd = 1.5) +
labs(y = "GPA", x = "satv Percentile",
title = "satv Effect plot with 95% CI and PI") +
theme_bw()

dm1 <- tibble(satv = rep(48.93, 50), satm = seq(from = 29, to = 77, length.out = 50))
mm1 <- as_tibble(predict(gpa1, newdata = dm1, interval = "confidence"))
pm1 <- as_tibble(predict(gpa1, newdata = dm1, interval = "prediction"))
mres_GPA_m <- bind_cols(dm1, mm1, pm1 %>% select(-fit))
#Rename CI and PI limits to have more explicit column names:
mres_GPA_m <- mres_GPA_m %>% rename(lwr_CI = lwr...4, upr_CI = upr...5,
lwr_PI = lwr...6, upr_PI = upr...7)
m1 <- mres_GPA_m %>% ggplot() +
geom_line(aes(x = satm, y = fit), lwd = 1) +
geom_ribbon(aes(x = satm, ymin = lwr_CI, ymax = upr_CI), alpha = .4,
fill = "beige", color = "darkred", lty = 2, lwd = 1) +
geom_ribbon(aes(x = satm, ymin = lwr_PI, ymax = upr_PI), alpha = .1,
fill = "gray80", color = "grey", lty = 3, lwd = 1.5) +
labs(y = "GPA", x = "satm Percentile",
title = "satm Effect plot with 95% CI and PI") +
theme_bw()
grid.arrange(v1, m1, ncol = 2)
```

#### a sidenote


```{r}
library(palmerpenguins)
dfp=penguins
dfp %>% ggplot(aes(x = flipper_length_mm,
                   y = bill_depth_mm)) +
  geom_point( size = 3,
             alpha = 0.8)+
theme_minimal()
```
```{r}
library(palmerpenguins)
dfp=penguins
dfp %>% ggplot(aes(x = flipper_length_mm,
                   y = bill_depth_mm)) +
  geom_point( size = 3,
             alpha = 0.8)+
     geom_smooth(method=lm)+
theme_minimal()
```

```{r}

dfp %>% ggplot(aes(x = flipper_length_mm,
                   y = bill_depth_mm,color=species)) +
  geom_point( size = 3,
             alpha = 0.8)+
     
theme_minimal()
```

```{r}
dfp %>% ggplot(
                       aes(x = flipper_length_mm,
                           y = bill_depth_mm,color=species)) +
  geom_point( size = 3,
             alpha = 0.8)+
  #theme_minimal() +
  
     geom_smooth(method=lm)+
  theme_minimal()

```

```{r}
library(palmerpenguins)
dfp=penguins
dfp %>% ggplot(
                       aes(x = body_mass_g,
                           y = bill_depth_mm,color=species)) +
  geom_point( size = 3,
             alpha = 0.8) +
  #theme_minimal() +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguin size, Palmer Station LTER",
       subtitle = "Flipper length and body mass for Adelie, Chinstrap and Gentoo Penguins",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species") +
     geom_smooth(method=lm)+
  theme_minimal()

```

### 8.9 Different intercepts for different groups


Sat    and   gender.

```{r}
# Make 1,2 coded gender into factor GENDER
satgpa <- satgpa %>% mutate(GENDER = factor(gender))
# Make category names clear but note that level names might be wrong
levels(satgpa$GENDER) <- c("MALE", "FEMALE")
satgpa %>% ggplot(mapping = aes(x = satv, y = fygpa, color = GENDER, shape = GENDER)) +
geom_smooth(method = "lm") +
geom_point(alpha = 0.7) +
theme_bw() +
scale_color_viridis_d(end = 0.8, option = "plasma") +
labs(title = "Scatterplot of GPA vs satv by gender")
satgpa %>% ggplot(mapping = aes(x = satm, y = fygpa, color = GENDER, shape = GENDER)) +
geom_smooth(method = "lm") +
geom_point(alpha = 0.7) +
theme_bw() +
scale_color_viridis_d(end = 0.8, option = "inferno") +
labs(title = "Scatterplot of GPA vs satm by gender")
```

```{r}
SATGENDER1 <- lm(fygpa ~ satm + GENDER, data = satgpa) #Fit lm with satm and GENDER
summary(SATGENDER1)
```

#### pg 349

```{r}
plot(allEffects(SATGENDER1))
```

```{r}
confint(SATGENDER1)
```

## 8.10  Additive MLR with two groups: Headache


```{r}
library(heplots)
data(Headache)
Headache <- as_tibble(Headache)
Headache
```

```{r}
Headache <- Headache %>% mutate(treatment = factor(treatment),
                                treatment = fct_relevel(treatment, "Control"))
```



```{r}
# Make treatment a factor and Control the baseline category
Headache %>% ggplot(mapping = aes(x = du1, y = du2, color = treatment,
shape = treatment)) +
geom_smooth(method = "lm", se = F) +
geom_point(size = 2.5) +
theme_bw() +
scale_color_viridis_d(end = 0.85, option = "inferno") +
labs(title = "Scatterplot of Maximum DB tolerance before &
after treatment (by treatment)")
```

```{r}
head1 <- lm(du2 ~ du1 + treatment, data = Headache)
summary(head1)
```

```{r}
plot(allEffects(head1, residuals = T), grid = T)
```


```{r}
%par(mfrow = c(2,2), oma = c(0,0,3,0))
plot(head1, pch = 16, sub.caption = "")
title(main="Plot of diagnostics for additive model with du1 and
treatment for du2", outer=TRUE)
```


```{r}
Headache <- Headache %>% mutate(resids = residuals(head1),
fits = fitted(head1)
)

# my addition
head_augmented=augment(head1,Headache)


Headache %>% ggplot(mapping = aes(x = fits, y = resids,
color = treatment, shape = treatment)) +
geom_smooth(method = "lm", se = F) +
geom_point(size = 2.5) +
theme_bw() +
scale_color_viridis_d(end = 0.85, option = "inferno") +
labs(title = "Scatterplot of Residuals vs Fitted by Treatment Group") +
facet_grid(~ treatment)
```

```{r}
vif(head1)
```

```{r}
head1R <- lm(du2 ~ du1, data = Headache)
summary(head1R)
```

##  8.11  Different slopes and different intercepts

A new example provides motivation for the consideration of different slopes and intercepts. A study was
performed to address whether the relationship between nonverbal IQs and reading accuracy differs between
dyslexic and non-dyslexic students. Two groups of students were identified, one group of dyslexic students
was identified first (19 students) and then a group of gender and age similar student matches were identified
(25 students) for a total sample size of n = 44, provided in the dyslexic3 data set from the smdata package
[Merkle and Smithson, 2018].

```{r}
library(smdata)
data("dyslexic3")
dyslexic3 <- dyslexic3 %>% mutate(dys = factor(dys))
levels(dyslexic3$dys) <- c("no", "yes")
dyslexic3 %>% ggplot(mapping = aes(x = ziq, y = score, color = dys, shape = dys)) +
geom_smooth(method = "lm") +
geom_point(size = 2, alpha = 0.5) +
theme_bw() +
scale_color_viridis_d(end = 0.7, option = "plasma") +
labs(title = "Plot of IQ vs Reading by dyslexia status",
x = "Standardized nonverbal IQ scores",
y = "Reading score") +
facet_grid(cols = vars(dys))
```

```{r}
dyslexic3 %>% ggplot(mapping = aes(x = ziq, y = score, color = dys, shape = dys)) +
geom_smooth(method = "lm") +
geom_point(size = 2, alpha = 0.5) +
theme_bw() +
scale_color_viridis_d(end = 0.7, option = "plasma")
```


```{r}
dys_model <- lm(score ~ ziq * dys, data = dyslexic3)
summary(dys_model)
```

```{r}
par(mfrow = c(2,2), oma = c(0,0,2,0))
plot(dys_model, pch = 16, sub.caption = "")
title(main="Plot of diagnostics for Dyslexia Interaction model", outer=TRUE)
```

```{r}
dyslexic3 <- dyslexic3 %>% mutate(resids = residuals(dys_model),
fits = fitted(dys_model)
)
dyslexic3 %>% ggplot(mapping = aes(x = fits, y = resids, color = dys, shape = dys)) +
geom_smooth(method = "lm", se = F) +
geom_point(size = 2.5) +
theme_bw() +
scale_color_viridis_d(end = 0.7, option = "plasma") +
labs(title = "Scatterplot of Residuals vs Fitted by Group")
```

```{r}
plot(allEffects(dys_model), ci.style = "bands", multiline = T, lty = c(1,2), grid = T)
```

```{r}
plot(allEffects(dys_model, residuals = T), lty = c(1,2), grid = T)
```

#### additive only model (bad)

```{r}
dys_modelR <- lm(score ~ ziq + dys, data = dyslexic3)
summary(dys_modelR)
```

```{r}
plot(allEffects(dys_modelR, residuals = T))
```

```{r}
dys_modelR2 <- lm(score ~ dys, data = dyslexic3)
summary(dys_modelR2)
plot(allEffects(dys_modelR2, residuals = T), grid = T)
```


```{r}
head2 <- lm(du2 ~ du1 * treatment, data = Headache)
summary(head2)
```


```{r}
head2 <- lm(du2 ~ du1 * treatment, data = Headache)
summary(head2)
```

```{r}
plot(allEffects(head2, residuals = T), grid = T) #version 1
plot(allEffects(head2), multiline = T, ci.style = "bands", grid = T,
lty = c(1:4), lwd = 2) #version 2
```

## 8.12 F-tests

```{r}
Anova(head2)
```

```{r}
head1 <- lm(du2 ~ du1 + treatment, data = Headache)
Anova(head1)
```

```{r}
head0=lm(du2 ~ du1 , data = Headache)
Anova(head0)
```
```{r}
summary(head0)
```

## 8.13 

```{r}
library(MuMIn)
options(na.action = "na.fail") #Must run this code once to use dredge
snotel2R <- snotel_s %>% slice(-c(9,22))
m6 <- lm(Snow.Depth ~ Elevation + Min.Temp + Max.Temp, data = snotel2R)

dredge(m6, rank = "AIC", extra = c("R^2",
                 "adjR^2",
                 myadjRsq = function(x) summary(x)$adj.r.squared
                 ))
```

