---
title: "Day 9 and 10 (week 5) class"
subtitle: "Ch5 - contingency tables"
author: "Joe Skufca"
date:  "2024-02-11"
output: html_notebook
---

Working through chapter 5 of *Greenwood*.

## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)

library(readxl) #for data ease
library(mosaic)
library(yarrr)
```
ANOVA chapter added four new packages:
```{r}
library(effects)
library(car)
library(catstats)
library(tabplot)
```

And for chapter 5, we add:

```{r}
library(vcd)  #visualizing categorical data
library(RColorBrewer) # colors
library(poLCA) # for a dataset
library(survey) # we will use for a dataset, but supports analysis of survey data
```

## Sec 5.1

```{r}
data(Arthritis)
 #Double-blind clinical trial with treatment and control groups

Arthritis <- Arthritis  %>% as_tibble()
# Homogeneity example

# Options needed to (sometimes) prevent errors on PC
# options(ffbatchbytes = 1024ˆ2 * 128); options(ffmaxbytes = 1024ˆ2 * 128 * 32)
tableplot(Arthritis, select = c(Treatment, Improved, Sex, Age), pals = list("BrBG"),
sample = F, colorNA_num = "orange", numMode = "MB-ML")
```

```{r}
dft=Arthritis %>% tabyl(Treatment,Improved)
dft
```

```{r}
tally(Treatment~Improved,data=Arthritis, margins=TRUE)
```
```{r}
tally(~Treatment+Improved,data=Arthritis, margins=TRUE)
```

```{r}
par(mai = c(1.5,1.5,0.82,0.42), #Adds extra space to bottom and left margin,
las = 1, #Rotates text labels, optional code
mgp = c(6,1,0)) #Adds space to labels, order is axis label, tick label, tick mark
plot(Improved ~ Treatment, data = Arthritis,
main = "Stacked Bar Chart of Arthritis Data")
```

## 5.2 Homogeneity test hypotheses

## 5.3 Independence test

```{r}
data(election, package = "poLCA")
election <- as_tibble(election) 
# Subset variables and remove missing values
election2 <- election %>% 
     dplyr::select(PARTY, VOTE3) %>%
mutate(VOTEF = factor(VOTE3)) %>%
drop_na()
levels(election2$VOTEF) <- c("Gore", "Bush", "Other") #Replace 1,2,3 with meaningful names
levels(election2$VOTEF) #Check new names of levels in VOTEF
```

```{r}
electable <- tally(~ PARTY+VOTEF, data = election2) #Contingency table
electable
par(las = 1) # horizontal labels
mosaicplot(electable, main = "Mosaic plot of observed results")
electable <- tally(~ VOTEF+PARTY, data = election2) #Contingency table
electable
par(las = 1) # horizontal labels
mosaicplot(electable, main = "Mosaic plot of observed results")
```

```{r}
par(las = 1) # horizontal labels
mosaicplot(electable, main = "Mosaic plot of observed results")
```

## 5.4 Models

## 5.5  Permutation test

```{r}
Arthtable <- tally(~ Treatment + Improved, data = Arthritis )

m=.75

m*Arthtable
chisq.test(m*Arthtable)
```

```{r}
c=chisq.test(10*Arthtable)
```

```{r}
c$expected
c$residuals^2
c$stdres^2
```

Maybe not reliable p-value.

```{r}
c
```
A permuted dataset:

```{r}
Arthperm <- Arthritis
Arthperm <- Arthperm %>% mutate(PermTreatment = shuffle(Treatment))
plot(Improved ~ PermTreatment, data = Arthperm,
main = "Stacked Bar Chart of Permuted Arthritis Data")
Arthpermtable <- tally(~ PermTreatment + Improved, data = Arthperm)
Arthpermtable
```
### A permutation test

```{r}
Tobs=c$statistic
```


```{r}
B=10000
Tstar=replicate(B,{
     chisq.test(tally(~ shuffle(Treatment) + Improved,
                      data = Arthritis))$statistic
})

pdata(Tstar,Tobs,lower.tail=FALSE)[[1]]


```

```{r}
tibble(Tstar) %>% ggplot(aes(x = Tstar)) +
geom_histogram(aes(y = ..ncount..), bins = 20, col = 1, fill = "khaki") +
geom_density(aes(y = ..scaled..)) +
theme_bw() +
labs(y = "Density") +
geom_vline(xintercept = Tobs, col = "red", lwd = 2) +
stat_bin(aes(y = ..ncount.., label = ..count..), bins = 20,
geom = "text", vjust = -0.75)
```

## 5.6 chi-square

If cell counts sufficiently large:

```{r}
chisq.test(Arthtable)$expected
```

```{r}
chisq.test(Arthtable)
```

### 5.7

```{r}
chisq.test(Arthtable)$residuals
mosaicplot(5*Arthtable, shade = T)
```

### 5.9


```{r}
election <- election %>% mutate(VOTEF = factor(VOTE3),
PARTY = factor(PARTY),
EDUC = factor(EDUC),
GENDER = factor(GENDER)
)
levels(election$VOTEF) <- c("Gore","Bush","Other")
# (Possibly) required options to avoid error when running on a PC,
# should have no impact on other platforms
# options(ffbatchbytes = 1024ˆ2 * 128); options(ffmaxbytes = 1024ˆ2 * 128 * 32)
tableplot(election, select = c(VOTEF, PARTY, EDUC, GENDER), pals = list("BrBG"),
sample = F)
```



```{r}
election2 <- election %>%
select(VOTEF, PARTY, EDUC, GENDER) %>%
drop_na()
tableplot(election2, select = c(VOTEF, PARTY, EDUC, GENDER), sort = 1,
pals = list("BrBG"), sample = F)
```

### 5.10 cheating and lying

```{r}
library(poLCA)
data(cheating) #Survey of students
cheating <- as_tibble(cheating)
cheating <- cheating %>% mutate(LIEEXAM = factor(LIEEXAM),
LIEPAPER = factor(LIEPAPER),
FRAUD = factor(FRAUD),
COPYEXAM = factor(COPYEXAM),
GPA = factor(GPA)
)
tableplot(cheating, sort = GPA, pals = list("BrBG"))
```

```{r}
cheating <- cheating %>% mutate(liar = interaction(LIEEXAM, LIEPAPER),
copier = interaction(FRAUD, COPYEXAM)
)
levels(cheating$liar) <- c("None", "ExamLie", "PaperLie", "LieBoth")
levels(cheating$copier) <- c("None", "PaperCheat", "ExamCheat", "PaperExamCheat")
tableplot(cheating, sort = liar, select = c(liar, copier), pals = list("BrBG"))
```

```{r}
cheatlietable <- tally(~ liar + copier, data = cheating)
cheatlietable
```

```{r}
# Collapse the middle categories of both variables by making both have the same level name:
cheating <- cheating %>% mutate(liar2 = liar,
copier2 = copier
)
levels(cheating$liar2) <- c("None", "ExamorPaper", "ExamorPaper", "LieBoth")
levels(cheating$copier2) <- c("None", "ExamorPaper", "ExamorPaper", "CopyBoth")
tableplot(cheating, sort = liar2, select = c(liar2, copier2), pals = list("BrBG"))
```





```{r}
cheatlietable <- tally(~ liar2 + copier2, data = cheating)
cheatlietable
```

### 5.11 Stratified random sample

```{r}
library(survey)
data(api)
apistrat <- as_tibble(apistrat)
apipop <- as_tibble(apipop)
tally(~ stype, data = apipop) #Population counts
```

```{r}
tally(~ stype, data = apistrat) #Sample counts
##
```

```{r}
m1 <- lm(growth ~ stype, data = apistrat)
library(car)
Anova(m1)
```

```{r}
plot(m1, which = 2, pch = 16)
```

```{r}
shapiro.test(apistrat$growth)
```

```{r}
favstats(~ growth, data = apistrat)
```


