---
title: "Day 2 class"
subtitle: "Refresher on basic Rstudio operations"
author: "Joe Skufca"
date:  "2024-01-17"
output: html_notebook
---

## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)
```

I am pursuing materials from Greenwoods book.


As we progress, we will add other packages that are needed.   For example, I know I intend to read an excel file, so:

```{r}
library(readxl)
```

Additionally, Greenwood uses the mosaic package:

```{r}
library(mosaic)
```

## Data load

Let's play with the treadmill dataset:

```{r}
treadmill=read_excel(here("data","treadmill.xls")) %>% 
     clean_names()
```

From page 11, we use the `favstats` function:

```{r}
treadmill %>% favstats(~run_time,data=.)
```

From page 18:

```{r}
ggplot(data = treadmill, mapping = aes(x = run_time)) +
geom_histogram(fill = "grey", bins = 8) +
geom_boxplot(color = "tomato") +
geom_rug(color = "skyblue", sides = "b", aes(y = 0), position = "jitter") +
theme_light()
```

# Practice problems

1.1 (Creating rmarkdown and loading data) is already accomplished.

### 1.2

Read in the treadmill data set discussed previously and find the mean and SD of the Ages (Age variable)
and Body Weights (BodyWeight variable). In studies involving human subjects, it is common to report a
summary of characteristics of the subjects. Why does this matter? Think about how your interpretation of
any study of the fitness of subjects would change if the mean age (same spread) had been 20 years older or
35 years younger.


```{r}
mean(treadmill$age)
sd(treadmill$age)

mean(treadmill$body_weight)
sd(treadmill$body_weight)
```



### 1.3. 

How does knowing about the distribution of results for Age and BodyWeight help you understand the
results for the Run Times discussed previously?

```{r}
ggplot(data = treadmill, mapping = aes(x = age)) +
geom_histogram(fill = "grey", bins = 8) +
geom_boxplot(color = "tomato") +
geom_rug(color = "skyblue", sides = "b", aes(y = 0), position = "jitter") +
theme_light()


ggplot(data = treadmill, mapping = aes(x = body_weight)) +
geom_histogram(fill = "grey", bins = 8) +
geom_boxplot(color = "tomato") +
geom_rug(color = "skyblue", sides = "b", aes(y = 0), position = "jitter") +
theme_light()
```

### 1.5.
The weight responses are in kilograms and you might prefer to see them in pounds. The conversion is
lbs = 2.205*kgs. Create a new variable in the treadmill tibble called BWlb using this code:


```{r}
treadmill$bwlb=2.205*treadmill$body_weight

mean(treadmill$bwlb)
sd(treadmill$bwlb)

```

### 1.6. 
Make histograms and boxplots of the original BodyWeight and new BWlb variables, both using base
R plots and using ggplot2. Discuss aspects of the distributions that changed and those that remained the
same with the transformation from kilograms to pounds. What does this tell you about changing the units of
a variable in terms of its distribution?


```{r}
ggplot(data = treadmill, mapping = aes(x = bwlb)) +
geom_histogram(fill = "grey", bins = 8) +
geom_boxplot(color = "tomato") +
geom_rug(color = "skyblue", sides = "b", aes(y = 0), position = "jitter") +
theme_light()


```