---
title: "Day 3 class"
subtitle: "Re-Introduction"
author: "Joe Skufca"
date:  "2024-01-17"
output: html_notebook
---

Working through chapter 2 of *Greenwood*.

## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)

library(readxl) #for data ease
library(mosaic)
library(yarrr)
```



## Data load

`readr` can (often) read directly from a URL if the URL is simply a data file.

```{r}
dd <- read_csv("http://www.math.montana.edu/courses/s217/documents/Walker2014_mod.csv") %>% clean_names()
```

```{r}
dd1=dd %>%
     mutate(across(where(is.character),factor))
```

```{r}
summary(dd1)
```

Take the time to label variables and the plot.   It will pay divididends in the future.

```{r}
dd %>% ggplot(mapping = aes(x = distance)) +
     geom_histogram(bins = 20, fill = "grey") +
     geom_rug(alpha = 0.1) +
     geom_boxplot(color = "tomato", width = 30) +
# width used to scale boxplot to make it more visible
     theme_bw() +
     labs(title = "Plot of Passing Distances",
          x = "Distance(cm)",y="Count")
```


Note that we now use "density" for the histogram, and add a density curve.

```{r}
dd1 %>% ggplot(mapping = aes(x = distance)) +
     geom_histogram(bins = 20, fill = "grey",aes(y = ..density..)) +
     geom_density(fill="purple",alpha=.1)+
     geom_rug(alpha = 0.1) +
# width used to scale boxplot to make it more visible
     theme_bw() +
     labs(title = "Plot of Passing Distances",
          x = "Distance(cm)",y="Count")
```

```{r}
library(yarrr)
pirateplot(distance ~ condition, data = dd1, inf.method = "ci", inf.disp = "line")
```
## subsetting data

```{r}
ddsub <- dd1 %>%
filter(condition %in% c("casual","commute")) %>%
select(distance, condition)

```

```{r}
ddsub <- ddsub %>% mutate(condition = factor(condition))
```

```{r}
mean(distance ~ condition, data = ddsub)
```

### exploring a small sample

```{r}
set.seed(9432)
s1 <- sample(ddsub %>% filter(condition %in% "commute"), size = 15)
s2 <- sample(ddsub %>% filter(condition %in% "casual"), size = 15)
dsample <- rbind(s1, s2)
mean(distance ~ condition, data = dsample)

```
```{r}
perm1 <- dsample %>%
select(distance, condition) %>%
mutate(permuted_condition = shuffle(condition))
# To force the tibble to print out all rows in data set -- not used often
#data.frame(Perm1)

```


```{r}
mean(distance ~ permuted_condition, data = perm1)
```

```{r}
lm1 <- lm(distance ~ condition, data = dsample)
summary(lm1)

```

```{r}
coef(lm1)[2]
```

```{r}
lmP <- lm(distance ~ permuted_condition, data = perm1)
coef(lmP)[2]

```

```{r}
B <- 1000

Tstar <- matrix(NA, nrow = B)
#Tstar

for (b in (1:B)){
lmP <- lm(distance ~ shuffle(condition), data = dsample)
Tstar[b] <- coef(lmP)[2]
#print(Ts)
}

```

