---
title: "Day 5 and 6 class"
subtitle: "ANOVA"
author: "Joe Skufca"
date:  "2024-01-17"
output: html_notebook
---

Working through chapter 3 of *Greenwood*.

## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)

library(readxl) #for data ease
library(mosaic)
library(yarrr)
```



## Data load

`readr` can (often) read directly from a URL if the URL is simply a data file.

We clean the names and convert non-numeric variables to factor.

```{r}
dd <- read_csv("http://www.math.montana.edu/courses/s217/documents/Walker2014_mod.csv") %>% 
     clean_names() %>%
     mutate(across(where(is.character),factor))
```

Take the time to label variables and the plot.   It will pay divididends in the future.

```{r}
dd %>% ggplot(mapping = aes(x = distance)) +
     geom_histogram(bins = 20, fill = "grey") +
     geom_rug(alpha = 0.1) +
     geom_boxplot(color = "tomato", width = 30) +
# width used to scale boxplot to make it more visible
     theme_bw() +
     labs(title = "Plot of Passing Distances",
          x = "Distance(cm)",y="Count")
```


Our pirate plot where I have included an option to explore with less data.

```{r}
pirateplot(distance ~ condition, 
           data = dd |> slice_sample(prop=1,by=condition), 
           inf.method = "ci", inf.disp = "line")
```
### Section 3.2 ANOVA

```{r}
lm2 <- lm(distance ~ condition, data = dd)
summary(lm2)$coefficients
```

From above, some differences seem to have STRONG evidence supporting a difference.



Modeling with no constant.   Everything is referenced to mean of 0.

```{r}
lm1 <- lm(distance ~ condition - 1, data = dd)
summary(lm1)$coefficients
```
We see strong evidence that none of the group means was 0.

#### effects package

```{r}
summary(lm2)$coefficients
```


```{r}
library(effects)
plot(allEffects(lm2))
```

```{r}
lm3 <- lm(distance ~ 1, data = dd)
summary(lm3)$coefficients
```
## Sec 3.3 one-way ANOVA

```{r}
anova(lm2)
```


```{r}
Tobs <- anova(lm(distance ~ condition, data = dd))[1,2]
Tobs
```

Shuffle test:

```{r}
#need code here
```


```{r}
B <- 1000

Tstar =replicate(B, {
     anova(lm(distance ~ shuffle(condition), data = dd))[1,2]
}) 



```



```{r}
tibble(Tstar) %>% ggplot(aes(x = Tstar)) +
geom_histogram(aes(y = ..ncount..), bins = 20, col = 1, fill = "skyblue") +
geom_density(aes(y = ..scaled..)) +
theme_bw() +
labs(y = "Density") +
geom_vline(xintercept = Tobs, col = "red", lwd = 2) +
stat_bin(aes(y = ..ncount.., label = ..count..), bins = 20,
geom = "text", vjust = -0.75)
```

```{r}
pdata(Tstar, Tobs, lower.tail = F)[[1]]
```
##### parametric p-value (F statistic)

```{r}
anova(lm2)
```

## 3.4 ANOVA diagnostics

```{r}
par(mfrow = c(2,2))  # show plots in a 2x2 array
plot(lm2)
```

### 3.5 Toothgrowth

```{r}
tooth_growth=as_tibble(ToothGrowth) %>% clean_names()
```

```{r}
tally(~ supp+dose, data = tooth_growth) #Supplement Type (VC or OJ)
```

```{r}
tooth_growth2 <- tooth_growth %>% mutate(treat = interaction(supp, dose))
# New variable that combines supplement type and dosage
tally(~ treat, data = tooth_growth2)

```
```{r}
pirateplot(len ~ treat, data = tooth_growth2, inf.method = "ci", inf.disp = "line",
ylab = "Odontoblast Growth in microns", point.o = .7)
```


```{r}
m2 <- lm(len ~ treat, data = tooth_growth2)

plot(m2)

```


# Below here not fixed



```{r}
lmP <- lm(distance ~ permuted_condition, data = perm1)
coef(lmP)[2]

```

```{r}
B <- 1000

Tstar <- matrix(NA, nrow = B)
#Tstar

for (b in (1:B)){
lmP <- lm(distance ~ shuffle(condition), data = dsample)
Tstar[b] <- coef(lmP)[2]
#print(Ts)
}

```

