---
title: "Day 7 and 8 (week4) class"
subtitle: "Two-way ANOVA"
author: "Joe Skufca"
date:  "2024-02-05"
output: html_notebook
---

Working through chapter 4 of *Greenwood*.

## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)

library(readxl) #for data ease
library(mosaic)
library(yarrr)
```
ANOVA chapter uses three new packages:
```{r}
library(effects)
library(car)
library(catstats)
```



## Pretend data

```{r}
pt <- read_csv("http://www.math.montana.edu/courses/s217/documents/pt.csv") %>%
     mutate(across(brand:drops,factor)) %>%
     clean_names() %>%
     select(-dropsf)
```

## 4.2



```{r}
tally(brand ~ drops, data = pt)
```

```{r}
favstats(responses ~ brand + drops, data = pt)
```

### pg 137

```{r}
set.seed(12)
pirateplot(responses ~ brand * drops, data = pt, xlab = "Drops", ylab = "Time",
inf.method = "ci", inf.disp = "line", theme = 2, point.o = 1)
```

```{r}
intplot(responses ~ brand * drops, data = pt)
```

```{r}
intplot(responses ~ drops * brand, data = pt)
```

### pg 139
```{r}
intplotarray(responses ~ drops * brand, data = pt)
```


## Sec 4.3  


```{r}
m1 <- lm(responses ~ brand * drops, data = pt)
anova(m1)
```

```{r}
plot(allEffects(m1), grid = T, multiline = T, lty = c(1:2), ci.style = "bars")
```


```{r}
m2 <- lm(responses ~ brand + drops, data = pt)
anova(m2)
```

```{r}
Anova(m2)
```


#### Illustrate with subsample

```{r}
pt_sub=pt |> slice_sample(prop=.9)
m2_1 <- lm(responses ~ brand + drops, data = pt_sub)
m2_2 <- lm(responses ~ drops + brand, data = pt_sub)
anova(m2_1)
anova(m2_2)
```

```{r}
pt_sub=pt |> slice_sample(prop=.9)
m2_1 <- lm(responses ~ brand + drops, data = pt_sub)
m2_2 <- lm(responses ~ drops+brand, data = pt_sub)
Anova(m2_1)
Anova(m2_2)
```
```{r}

#pt_sub=pt |> slice_sample(prop=.9)
m2_1f <- lm(responses ~ brand + drops, data = pt)
m2_2f <- lm(responses ~ drops+brand, data = pt)
anova(m2_1f)
anova(m2_2f)

```

### pg 146

```{r}
summary(m2)
```

```{r}
plot(allEffects(m2, residuals = T))
```


## 4.4  Guinea pig

```{r}
ToothGrowth=ToothGrowth |>
     as_tibble() |>
     clean_names()
```


Look at the data, thinking about interractions:

```{r}
par(mfrow = c(1,2))

pirateplot(len ~ supp * dose, data = ToothGrowth, ylim = c(0,35),
main = "Pirate-plot", xlab = "Dosage", ylab = "Odontoblast Growth",
inf.method = "ci", inf.disp = "line", theme = 2)

intplot(len ~ supp * dose, data = ToothGrowth, col = c(1,2),
main = "Interaction Plot", ylim = c(0,35))
```
Balanced?

```{r}
tabyl(ToothGrowth,supp,dose)
```



Test for an interraction:

```{r}
TG1 <- lm(len ~ supp * dose, data = ToothGrowth)
Anova(TG1)

```

```{r}
ToothGrowth <- ToothGrowth %>% mutate(dosef = factor(dose))
TG2 <- lm(len ~ supp * dosef, data = ToothGrowth)
Anova(TG2)
```

Keep the interaction - and focus on the interaction.

```{r}
par(mfrow = c(2,2))
plot(TG2, pch = 16)
```
Equal variances?

```{r}
plot(allEffects(TG2, residuals = T, x.var = "dosef"))
```

```{r}
summary(TG2)$coefficients
```

```{r}
tg2 <- aov(len ~ supp * dosef, data = ToothGrowth)
```


```{r}
summary(TG2)
summary(tg2)
```

```{r}
#summary(TG2)$coefficients
tg2$coefficients
```

## 4.5 The Psychology of Debt



```{r}
library(faraway)
data(debt)

debt <- debt |>
     as_tibble() %>% 
     mutate(incomegp = factor(incomegp),
                                cigbuy = factor(cigbuy)) 
     
debtc <- debt %>% drop_na()

debtRc <- debt %>% drop_na(incomegp, cigbuy, prodebt)
```


```{r}


pirateplot(prodebt ~ cigbuy * incomegp, data = debtRc, 
main = "Pirate-plot", 
inf.method = "ci", inf.disp = "line", theme = 2)

```

```{r}


intplotarray(prodebt ~ cigbuy * incomegp, data = debtRc, col = c(1,3,4,5,6), lwd = 2)
```

```{r}
tally(cigbuy ~ incomegp, data = debtRc)
```


```{r}
debt1 <- lm(prodebt ~ incomegp * cigbuy, data = debtRc)
Anova(debt1)
```

```{r}
par(mfrow = c(2,2))
plot(debt1, pch = 16)
```

#### Continuing without interaction







```{r}
debt1r <- lm(prodebt ~ incomegp + cigbuy, data = debtRc)
par(mfrow = c(2,2))
plot(debt1r, pch = 16)
```

```{r}
plot(allEffects(debt1r, residuals = T))
```

```{r}
Anova(debt1r)
```

```{r}
plot(allEffects(debt1r))
```

```{r}
summary(debt1r)$coefficients
```

Should we do Tukey?

## 4.6  Un-replicated designs

```{r}
ptR <- read_csv("http://www.math.montana.edu/courses/s217/documents/ptR.csv")
ptR <- ptR %>% mutate(dropsf = factor(drops),
brand = factor(brand))
ptR
```

```{r}
intplot(responses ~ brand * dropsf, data = ptR, lwd = 2)
```

### Pg 165

```{r}
anova(lm(responses ~ dropsf * brand, data = ptR))
```

```{r}
norep1 <- lm(responses ~ dropsf + brand, data = ptR)
Anova(norep1)
```

```{r}
plot(allEffects(norep1))
```

```{r}
#remotes::install_github("mtennekes/tabplot")
```



```{r}
dd <- read_csv("http://www.math.montana.edu/courses/s217/documents/Walker2014_mod.csv")
dd <- dd %>% mutate(Condition = factor(Condition),
Condition2 = reorder(Condition, Distance, FUN = mean),
Shirt = factor(Shirt),
Helmet = factor(Helmet),
Pants = factor(Pants),
Gloves = factor(Gloves),
ReflectClips = factor(ReflectClips),
Backpack = factor(Backpack)
)
```
```{r}
library(tabplot)
library(RColorBrewer)
# Options (sometimes) needed to prevent errors on PC
# options(ffbatchbytes = 1024ˆ2 * 128); options(ffmaxbytes = 1024ˆ2 * 128 * 32)
tableplot(dd[,-1], sort = Condition2, pals = list("BrBG"), sample = F,
colorNA_num = "pink", numMode = "MB-ML")
```


```{r}
overtake_int <- lm(Distance ~ Helmet * ReflectClips, data = dd)
summary(overtake_int)
```

```{r}
tally(Helmet ~ ReflectClips, data = dd)
```

```{r}
aov(Distance ~ Helmet * ReflectClips, data = dd) 
```

