---
title: "Logistic Regression"
subtitle: "Part 1"
author: "Joe Skufca"
date: "2024-04-03"
output:
  html_notebook: default
---

This work take from Ch 6 of *Introduction to Regression Methods for Public Health Using R* by Nahhas.

https://www.bookdown.org/rwnahhas/RMPH/


Supporting material is from Ch 6 of *Beyond Multiple Linear Regression* by Roback and Legler

https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html

Specifically, will will tackle the guided homework problem on elephant mating.



## Packages

Recommended from the book:

```{r}
# Packages required for Chapter 4
library(gridExtra)  # grid graphics
library(knitr)   #knitting documents
library(kableExtra) # pretty tables
library(mosaic) # stats teaching utilities
library(xtable) # html tables
library(pscl) #poly sci computational lab
library(multcomp) 
library(pander) #pandoc writing
library(MASS) #modern applied statistics book
library(tidyverse)

```

Adding my own favorites.

```{r}

library(janitor)
library(here)
library(broom)

theme_set(theme_minimal())
```

# Some artificial data about hearing

The logit and its inverse are available in base R, but under differengt names, so
I redefine here for readability.

```{r}
logit <- qlogis
invlogit <- plogis
```


```{r}
set.seed(1)
n=40

power=runif(n,0,40)
age=sample(19:30,n,replace=TRUE)

y=-4.3 +.25*power
r=rbinom(n,1,invlogit(y))
df1=tibble(power,age,r)


#write_csv(df1,here::here("data","hearing.csv"))
```

## visualization

```{r}
df1 %>% ggplot(aes(power,r)) +
     geom_point()
```

```{r}
df1 %>% ggplot(aes(power,r)) +
     geom_point() + 
     geom_smooth(method=lm,color="red",se=FALSE)
```

```{r}
df1 %>% ggplot(aes(power,r)) +
     geom_point() + 
     geom_smooth(method=lm,color="red",se=FALSE)+
     geom_smooth(method = "glm", 
                 method.args = list(family = "binomial"),
                 se = FALSE) 
```

```{r}
df1 %>% ggplot(aes(power,r)) +
     geom_point() + 
     geom_smooth(method=lm,color="red",se=FALSE)+
     geom_smooth(method = "glm", 
                 method.args = list(family = "binomial"),
                 se = FALSE) +
     geom_smooth(method = "glm", 
                 method.args = list(family = binomial(link="probit")),
                 se = FALSE,color="green") 
```

## Now the regression analysis

```{r}
m1=glm(r~power,data=df1,family=binomial)

summary(m1)
```

What if we add `age` to the regression:

```{r}

m2=glm(r~power+age,data=df1,family=binomial)

summary(m2)

```


# A second dataset on hearing

Let's generate a dataset where the process includes a lurking variable:


```{r}
set.seed(1)
n=40

power=runif(n,0,40)
age=sample(19:30,n,replace=TRUE)
first_last=sample(c("first","last"),n,replace=TRUE)

y2=-4.3 +.25*power+4.5*(first_last=="last")
r2=rbinom(n,1,invlogit(y2))
df2=tibble(power,age,r2)



```




```{r}
m3=glm(r2~power,data=df2,family=binomial)

summary(m3)
```


```{r}
df2 %>% ggplot(aes(power,r)) +
     geom_point() + 
     geom_smooth(method=lm,color="red",se=FALSE)+
     geom_smooth(method = "glm", 
                 method.args = list(family = "binomial"),
                 se = FALSE) 
```