---
title: "Repeated Measures ANOVA"
subtitle: "Teaching lecture"
author: "Joe Skufca"
date:  "2024-04-15d"
output: html_notebook
---

* Revisiting a bit chapter 3 of *Greenwood*
* Background from https://www.sciencedirect.com/science/article/pii/S0022480407002569#sec1.4
* https://statistics.laerd.com/statistical-guides/repeated-measures-anova-statistical-guide-2.php
* Code based examples from https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/
* Code based https://www.r-bloggers.com/2021/04/repeated-measures-of-anova-in-r-complete-tutorial/


## Packages that we will typically load


Here are three of my favorites that I always load.

```{r}
library(tidyverse)
library(janitor)
library(here)

library(readxl) #for data ease
library(mosaic)
library(yarrr)
library(car)

theme_set(theme_minimal())
```



## A simulation with two groups

Let's try to better understand "repeated measures" through *simulated data* of
a type you are familiar.


### Some artificial data:

```{r}
set.seed(42)
n=20
x=rnorm(n,75,8)
y=x + rnorm(n,-1,.6)

mean(x)
mean(y)
x-y

```

```{r}
df1=tibble(score=c(x,y) %>% round(),sec=rep(c(1,2),each=n))
df1
```

```{r}
pirateplot(score~sec, 
           data = df1 , 
           inf.method = "ci", inf.disp = "line")

```
```{r}
m1=lm(score~sec,data=df1)
Anova(m1)
```

```{r}
t.test(score~sec,data=df1)
```

### Same data, but different story

```{r}
df2=tibble(t1=round(x),t2=round(y), tdiff=t2-t1)
df2
```

```{r}
t.test(df2$t1,df2$t2)
```

```{r}
t.test(df2$t1,df2$t2,paired=TRUE)
```
```{r}
df21=df2 %>% 
     rowid_to_column(var="stud") %>%
     pivot_longer(t1:t2,names_to = "test")
     
     
     
df21 %>%     
     ggplot(aes(x = test, y = value, group = stud)) + 
geom_line() +geom_point()


```

## Repeated measures in R

```{r}
library(ggpubr)
library(rstatix)
```


```{r}
data("selfesteem", package = "datarium")
head(selfesteem, 3)
```

```{r}
# Gather columns t1, t2 and t3 into long format
# Convert id and time into factor variables
selfesteem_a <- selfesteem %>%
  gather(key = "time", value = "score", t1, t2, t3) %>%
  rstatix::convert_as_factor(id, time)
head(selfesteem_a, 3)
```

```{r}
selfesteem_b = selfesteem %>% 
     pivot_longer(t1:t3,names_to = "time",values_to = "score") %>%
     convert_as_factor(id,time)
```


```{r}
selfesteem_b %>%
  group_by(time) %>%
  get_summary_stats(score, type = "mean_sd")
```

```{r}
bxp=selfesteem_b %>%
    ggplot(aes(time,score)) +
     geom_boxplot()+
     geom_point()

bxp
```
#### check for outliers

```{r}
selfesteem_b %>%
  group_by(time) %>%
  identify_outliers(score)
```

#### Check for normality




```{r}
selfesteem_b %>%
  group_by(time) %>%
  shapiro_test(score)
```

Note that, if your sample size is greater than 50, the normal QQ plot is preferred because at larger sample sizes the Shapiro-Wilk test becomes very sensitive even to a minor deviation from normality.

```{r}
selfesteem_b %>%
     ggqqplot("score",facet.by = "time")
  
```



#### Computation

`anova_test` from rstatix will do the computation for us.



```{r}
res.aov <- anova_test(data = selfesteem_b, dv = score, wid = id, within = time)
get_anova_table(res.aov)

```

#### post-hoc comparisons

```{r}
# pairwise comparisons
pwc <- selfesteem_b %>%
  pairwise_t_test(
    score ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc
```


#### Visualize

```{r}
# Visualization: box plots with p-values
pwc <- pwc %>% add_xy_position(x = "time")
bxp + 
  stat_pvalue_manual(pwc) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
  )
```




### revisit data frame 2

```{r}
res.aov <- anova_test(data = df21, dv = value, wid = stud, within = test)
get_anova_table(res.aov)
```

##### outliers, normality, sphericity

```{r}
df21 %>%
  group_by(test) %>%
  identify_outliers(value)

df21 %>%
  group_by(test) %>%
  shapiro_test(value)

```

## Two-way repeated measures

```{r}
# Wide format
set.seed(123)
data("selfesteem2", package = "datarium")
```




What do we do first?






Make the data long:



```{r}
selfesteem2a =selfesteem2 %>%
     pivot_longer(t1:t3,names_to="time",values_to = "score") %>%
     convert_as_factor(id,time)

selfesteem2a
```


```{r}
bxp <- ggboxplot(
  selfesteem2a, x = "time", y = "score",
  color = "treatment", palette = "jco"
  )
bxp
```

```{r}
selfesteem2a %>%
  group_by(treatment, time) %>%
  identify_outliers(score)
```


```{r}
selfesteem2a %>%
  group_by(treatment, time) %>%
  shapiro_test(score)
```



```{r}
ggqqplot(selfesteem2a, "score", ggtheme = theme_bw()) +
  facet_grid(time ~ treatment, labeller = "label_both")
```

```{r}
res.aov <- anova_test(
  data = selfesteem2a, 
  dv = score, 
  wid = id,
  within = c(treatment, time)
  )
get_anova_table(res.aov)
```

