---
title: "Beyond linear regression"
subtitle: "Ch4 Homework"
author: "Joe Skufca"
date: "2024-03-28"
output:
  html_notebook: default
---

This work take from Ch 4 of *Beyond Multiple Linear Regression* by Roback and Legler

https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html

Specifically, will will tackle the guided homework problem on elephant mating.



## Packages

Recommended from the book:

```{r}
# Packages required for Chapter 4
library(gridExtra)  # grid graphics
library(knitr)   #knitting documents
library(kableExtra) # pretty tables
library(mosaic) # stats teaching utilities
library(xtable) # html tables
library(pscl) #poly sci computational lab
library(multcomp) 
library(pander) #pandoc writing
library(MASS) #modern applied statistics book
library(car)
library(tidyverse)

```

Adding my own favorites.

```{r}

library(janitor)
library(here)
library(broom)

theme_set(theme_minimal())
```

## load data

```{r}
#load("Data/nhanes1718_adult_exam_sub_rmph.Rdata")
load(here("data","nhanes1718_adult_exam_sub_rmph.Rdata"))

# Create dichotomized PHQ-9

# "PHQ-9 scores of 5, 10, 15, and 20 represented mild, moderate,
# moderately severe, and severe depression, respectively"
# (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1495268/)

nhanes <- nhanes_adult_exam_sub %>% 
  mutate(depression_mild = factor(phq9 >= 5,
                                  levels = c(F, T),
                                  labels = c("No", "Yes"))) %>%
     clean_names()
# Check
table(nhanes$phq9, nhanes$depression_mild, exclude = NULL)
tapply(nhanes$phq9, nhanes$depression_mild, range, na.rm=T)
```

## prob 4

2x2 table

```{r}
TAB=table(nhanes$paq635,nhanes$depression_mild)
TAB

(TAB[1,2]*TAB[2,1]) / (TAB[2,2]*TAB[1,1])
```

## prob 5

```{r}
TAB=table(nhanes$slq050,nhanes$depression_mild)
TAB

(TAB[1,2]*TAB[2,1]) / (TAB[2,2]*TAB[1,1])
```

## prob 6


```{r}
levels(nhanes$depression_mild)
```
No is the reference and glm models $p=P(Y="Yes")$.

## prob 7

Compute the odds ratio comparing the odds of mild depression (depression_mild) between those who do and those who do not answer “Yes” to the sleeping question (SLQ050) using logistic regression, as well as its 95% confidence interval. Assume depression_mild is the outcome.

```{r}
m1=glm(depression_mild ~paq635, data=nhanes, family=binomial)
summary(m1)
```

```{r}
exp(coef(m1))[-1]
```
```{r}
1/exp(coef(m1))[-1]
```

```{r}
1/exp(confint(m1))[-1, , drop=F]
```
## Prob 8
Do the odds of mild depression (depression_mild) differ between individuals of different income levels (income)? Test the global significance of income using a Type III Wald test and compute the OR, 95% CI, and p-value comparing each possible pair of levels.

```{r}
m2=glm(depression_mild ~income, data=nhanes, family=binomial)
summary(m2)


```

```{r}
car::Anova(m2,type=3,test.statistic = "Wald")
```

```{r}
exp(coef(m1))
```

```{r}
glht(m2,mcp(income="Tukey"))
```

## Prob 9

Is mild depression (depression_mild) associated with the number of days someone engages in vigorous recreational activities (PAQ655, “In a typical week, on how many days do you do vigorous-intensity sports, fitness or recreational activities?”)? What is the OR comparing mild depression between individuals who differ by 1 day in days of vigorous recreation? What about between those who differ by 5 days?

```{r}
m3=glm(depression_mild ~ paq655, data=nhanes, family=binomial)
summary(m3)

exp(coef(m3))[-1]

```

```{r}
nhanes %>% ggplot(aes(paq655,depression_mild))+
     geom_point()+
     geom_smooth(method = "glm", 
                 method.args = list(family = "binomial"),
                 se = FALSE)+
     geom_jitter()
```




## Prob 10

Is mild depression (depression_mild) significantly associated with trouble sleeping (SLQ050) after adjusting for age (RIDAGEYR), gender (RIAGENDR), income (income), and days someone engages in vigorous recreational activities (PAQ655)? Answer the question and report the AORs, 95% confidence intervals, and p-values. Also, interpret the AOR for trouble sleeping.

```{r}
#m4=glm(depression_mild ~ riagendr, data=nhanes, family=binomial)
m4=glm(depression_mild ~ slq050+ridageyr+riagendr+income+paq655, data=nhanes, family=binomial)
summary(m4)

exp(coef(m4))[-1]
exp(confint(m4))[-1]
```
```{r}
car::Anova(m4, type = 3, test.statistic = "Wald")
```

```{r}
S(m4)
```


## Prob 11
Create a forest plot to illustrate the AORs and their 95% CIs for the model from the previous Exercise. For each continuous predictor, plot the AOR corresponding to a difference in the predictor equal to its inter-quartile range (IQR).

```{r}
library(sjPlot)
#devtools::install_github("strengejacke/strengejacke")
```

```{r}
plot_model(m4)
```





