---
title: "Beyond linear regression"
subtitle: "Ch4 Homework"
author: "Joe Skufca"
date: "2024-03-28"
output:
  html_notebook: default
---

This work take from Ch 4 of *Beyond Multiple Linear Regression* by Roback and Legler

https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html

Specifically, will will tackle the guided homework problem on elephant mating.



## Packages

Recommended from the book:

```{r}
# Packages required for Chapter 4
library(gridExtra)  # grid graphics
library(knitr)   #knitting documents
library(kableExtra) # pretty tables
library(mosaic) # stats teaching utilities
library(xtable) # html tables
library(pscl) #poly sci computational lab
library(multcomp) 
library(pander) #pandoc writing
library(MASS) #modern applied statistics book
library(car)
library(tidyverse)

```

Adding my own favorites.

```{r}

library(janitor)
library(here)
library(broom)

theme_set(theme_minimal())
```

## load data

```{r}
#load("Data/nhanes1718_adult_exam_sub_rmph.Rdata")
load(here("data","nhanes1718_adult_exam_sub_rmph.Rdata"))

# Create dichotomized PHQ-9

# "PHQ-9 scores of 5, 10, 15, and 20 represented mild, moderate,
# moderately severe, and severe depression, respectively"
# (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1495268/)

nhanes <- nhanes_adult_exam_sub %>% 
  mutate(depression_mild = factor(phq9 >= 5,
                                  levels = c(F, T),
                                  labels = c("No", "Yes"))) %>%
     clean_names()
# Check
table(nhanes$phq9, nhanes$depression_mild, exclude = NULL)
tapply(nhanes$phq9, nhanes$depression_mild, range, na.rm=T)
```

## prob 4

2x2 table

```{r}
TAB=table(nhanes$paq635,nhanes$depression_mild)
TAB

(TAB[1,2]*TAB[2,1]) / (TAB[2,2]*TAB[1,1])
```

## prob 5

```{r}
TAB=table(nhanes$slq050,nhanes$depression_mild)
TAB

(TAB[1,2]*TAB[2,1]) / (TAB[2,2]*TAB[1,1])
```

## prob 6


```{r}
levels(nhanes$depression_mild)
```
No is the reference and glm models $p=P(Y="Yes")$.

## prob 7

Compute the odds ratio comparing the odds of mild depression (depression_mild) between those who do and those who do not answer “Yes” to the sleeping question (SLQ050) using logistic regression, as well as its 95% confidence interval. Assume depression_mild is the outcome.

```{r}
m1=glm(depression_mild ~slq050, data=nhanes, family=binomial)
summary(m1)
```

```{r}
exp(coef(m1))[-1]
```
```{r}
1/exp(coef(m1))[-1]
```

```{r}
1/exp(confint(m1))[-1, , drop=F]
```
## Prob 8
Do the odds of mild depression (depression_mild) differ between individuals of different income levels (income)? Test the global significance of income using a Type III Wald test and compute the OR, 95% CI, and p-value comparing each possible pair of levels.

```{r}
m2=glm(depression_mild ~income, data=nhanes, family=binomial)
summary(m2)


```

```{r}
car::Anova(m2,type=3,test.statistic = "Wald")
```

```{r}
exp(coef(m1))
```

```{r}
glht(m2,mcp(income="Tukey"))
```

## Prob 9

Is mild depression (depression_mild) associated with the number of days someone engages in vigorous recreational activities (PAQ655, “In a typical week, on how many days do you do vigorous-intensity sports, fitness or recreational activities?”)? What is the OR comparing mild depression between individuals who differ by 1 day in days of vigorous recreation? What about between those who differ by 5 days?






